---
title: "Cleaning and Wrangling Data"
author: "Maya Almaraz"
date: "2023-04-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Packages
```{r load packages, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyr)
library(readr)
```


# Introduction

About the data:
This data was downloaded from [this link](https://knb.ecoinformatics.org/view/df35b.304.2), on April 4, 2023.

## Read data
```{r}
catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
```


## Explore data

```{r explore, eval=FALSE, echo=FALSE}
colnames(catch_original)
head(catch_original)
summary(catch_original)
sort(unique(catch_original$Year))
unique(catch_original$Region)
View(catch_original)
```


# Clean and Reshape data

## Selecting and removing columns with `select()`
```{r select}
catch_data <- catch_original %>%
  select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)

head(catch_data)

catch_data <- catch_original %>% 
  select(-All, -notesRegCode)

catch_data <- select(catch_original, -All, -notesRegCode)

head(catch_data)
```


## Check column types
```{r checking}

glimpse(catch_data)
```

## Changing column content using `mutate()`

```{r changing data type}

catch_clean <- catch_data %>% 
  mutate(Chinook = as.numeric(Chinook))

head(catch_clean)
```

## Checking NA
```{r checking na}

i <- which(is.na(catch_clean$Chinook))

catch_data[i,]

```
## Fixing Chinook column

```{r}
catch_clean <- catch_data %>% 
  mutate(Chinook = if_else(Chinook == "I", 
                           "1",
                           Chinook), 
         Chinook = as.numeric(Chinook))

## Check
catch_clean[i,]
View(catch_data)
```


##Reshape data

From wide to long use `pivot_longer()`
```{r reshape}

catch_long <- catch_clean %>% 
  pivot_longer(cols = -c(Region, Year),
               names_to = "species",
               values_to = "catch")
head(catch_long)
```
From long to wide use `pivot_wider()`

```{r reshape wide}

catch_wide <- catch_long %>% 
  pivot_wider(names_from = species,
              values_from = catch)

head(catch_wide)

```
## Renaming column

```{r rename column}
catch_long <- catch_long %>% 
  rename(catch_thousands = catch)

colnames(catch_long)
```

## Add a column using `mutate()`

```{r adding column}

catch_long <- catch_long %>% 
  mutate(catch = catch_thousands*1000) %>% 
  select(-catch_thousands)

head(catch_long)
```
## Split-Apply-Combine workflow
We are going to use `group_by()` and `summarize()`

```{r split}

mean_region <- catch_long %>%
    group_by(Region) %>%
    summarize(mean_catch = mean(catch_thousands))

head(mean_region)

n_region <- catch_long %>% 
  group_by(Region) %>% 
  summarize(obs_n = n())

head(n_region)
  
```

## Filtering by rows with `filter()`

```{r filtering}

sse_catch <- catch_long %>% 
  filter(Region == "SSE")

head(sse_catch)

```
## All together

```{r all}

catch_original <- read_csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1", 
                               method = "libcurl"))


mean_region <- catch_original %>%
  select(-All, -notesRegCode) %>% 
  mutate(Chinook = ifelse(Chinook == "I", 1, Chinook)) %>% 
  mutate(Chinook = as.numeric(Chinook)) %>% 
  pivot_longer(-c(Region, Year), 
               names_to = "species", 
               values_to = "catch") %>%
  mutate(catch = catch*1000) %>% 
  group_by(Region) %>% 
  summarize(mean_catch = mean(catch))

head(mean_region)
```

