---
title: "Intro to Data Visualization"
author: "Maya Almaraz"
date: "2023-04-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages
```{r load packages, warning = FALSE, message = FALSE}

library(dplyr)
library(readr)
library(tidyr)
library(forcats)
library(ggplot2)
library(leaflet)
library(DT)
library(scales)

```

## Read data
```{r read data}

escapement <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3Af119a05b-bbe7-4aea-93c6-85434dcb1c5e")

```

Should add description of data and where it came from

## Explore data
```{r explore}

colnames(escapement)

glimpse(escapement)

range(escapement$sampleDate)

head(escapement)

unique(escapement$Species)

```

## Getting our data ready to plot

```{r cleaning}

annual_esc <- escapement %>% 
  separate(sampleDate, c("Year", "Month", "Day"), sep = "-") %>% 
  mutate(Year = as.numeric(Year)) %>% 
  group_by(Species, SASAP.Region, Year) %>% 
  summarize(escape = sum(DailyCount)) %>% 
  filter(Species %in% c("Chinook", "Sockeye", "Chum", "Coho", "Pink"))

unique(annual_esc$Species)

head(annual_esc)
```
## Plotting with `ggplot`

## Essential components
```{r essentials}
## Option 1
ggplot(data = annual_esc,
       aes(x = Species,
           y = escape))+ # plus sign adds layers
  geom_col() # to create column bar plots

## Option 2
ggplot(data = annual_esc)+
  geom_col(aes(x = Species, 
               y = escape))

# Option 3
ggplot()+
  geom_col(data = annual_esc,
           aes(x = Species, # aes defines mapping aesthetics
               y = escape)) 

```

## Different geoms

**Boxplot** 
```{r boxplot}

annual_esc %>% 
  filter(Year == 1974, # just plotting for one year of data
         Species %in% c("Chum", "Pink")) %>% # data you want to plot
  ggplot(aes(x = Species, # map these aesthetics
             y = escape))+
  geom_boxplot()

```
**Line and point**

```{r line and point}

annual_esc %>% 
  filter(Species == "Sockeye", # filter for sockeye
         SASAP.Region == "Bristol Bay") %>% # second filter
  ggplot(aes(x = Year,
             y = escape))+
  geom_line()+ # makes a line graph
  geom_point() # adds points to the line graph

```
### Customizing our plot
fill = blue inside the `aes()`

```{r custom blue}

ggplot(annual_esc,
       aes(x = Species,
           y = escape,
           fill = "blue"))+ # trying to make bars blue but failed
  geom_col()

# dont have a variable called fill so need to

```

fill = "blue" outside of `aes()`
```{r make blue}

ggplot(annual_esc,
       aes(x = Species,
           y = escape,))+ # data inside aes function
  geom_col(fill = "blue") # looks outside aes function

```

 Adding color by region
 
```{r color by region}

ggplot(annual_esc,
       aes(x = Species, 
           y = escape,
           fill = SASAP.Region)) +
    geom_col()

```
 Subset data
```{r subset}

annual_esc_2000 <- annual_esc %>% 
  filter(Year %in% c(2000:2016)) #filter to have the years we want

## quick check
unique(annual_esc_2000$Year)

```
 
 Plot with `facet_wrap()`
```{r faceting}

ggplot(annual_esc_2000,
       aes(x = Year, # define map aesthetics
           y = escape,
           color = Species))+ #color line
  geom_line()+
  geom_point()+
  facet_wrap(~SASAP.Region, #creates multiple plots at once
             scales = "free_y") # allows each plot to have their own scale
# ~ calls into datset with facet function (need to look into variable of data to create facet)

```
 
 ### Setting ggplot themes
 
 `theme_*()` function
 
```{r themes}

ggplot(annual_esc_2000,
       aes(x = Year, 
           y = escape,
           color = Species))+ 
  geom_line()+
  geom_point()+
  facet_wrap(~SASAP.Region, 
             scales = "free_y")
theme_bw()+ # black and white
  ylab("Escapement")+ # change y axis name
ggtitle("Annual Salmon Escapement by Region") # add title to graph

ggplot(annual_esc_2000,
       aes(x = Year, 
           y = escape,
           color = Species))+ 
  geom_line()+
  geom_point()+
  facet_wrap(~SASAP.Region, 
             scales = "free_y")+
theme_dark()+ # new theme
  ylab("Escapement")+ 
ggtitle("Annual Salmon Escapement by Region") 

ggplot(annual_esc_2000,
       aes(x = Year, 
           y = escape,
           color = Species))+ 
  geom_line()+
  geom_point()+
  facet_wrap(~SASAP.Region, 
             scales = "free_y")+
theme_light()+ # new theme
  ylab("Escapement")+ 
ggtitle("Annual Salmon Escapement by Region") 


```
 Removing legend title and moving legend 
```{r move legend}

ggplot(annual_esc_2000,
       aes(x = Year, 
           y = escape,
           color = Species))+ 
  geom_line()+
  geom_point()+
  facet_wrap(~SASAP.Region, 
             scales = "free_y")+
theme_bw()+ 
  ylab("Escapement")+ 
ggtitle("Annual Salmon Escapement by Region")+ # add another layer
  theme(legend.position = "bottom", # move legend to the bottom
        legend.title = element_blank()) # remove legend title

```
 ## Saving theme into an object
```{r saving theme}

my_theme <- theme_bw()+
  theme(legend.position = "bottom",
        legend.title = element_blank())

```
 
 Creating plot with saved theme
 
```{r use saved theme}

ggplot(annual_esc_2000,
       aes(x = Year,
           y = escape,
           color = Species)) +
    geom_line() +
    geom_point() +
    facet_wrap(~SASAP.Region,
                scales = "free_y") +
    ylab("Escapement") +
    ggtitle("Annual Salmon Escapement by Region") +
    my_theme

```
 Angle years to 45 degrees and x axis breaks by 2
```{r angle axis}

ggplot(annual_esc_2000,
       aes(x = Year,
           y = escape,
           color = Species)) +
    geom_line() +
    geom_point() +
    scale_x_continuous(breaks = seq(2000, 2016, 2)) + # axis in intervals of 2
    facet_wrap( ~ SASAP.Region,
                scales = "free_y") +
    ylab("Escapement") +
    ggtitle("Annual Salmon Escapement by Region") +
    my_theme +
    theme(axis.text.x = element_text(angle = 45, # angle axis
                                     vjust = 0.5))

```
 Smarter ticks using `scales()`
 
```{r smarter ticks}

annual_region_plot <- ggplot(annual_esc_2000,
       aes(x = Year,
           y = escape,
           color = Species)) +
    geom_line() +
    geom_point() +
    scale_x_continuous(breaks = seq(2000, 2016, 2)) +
    scale_y_continuous(labels = comma)+ #change axis to not scientific notation
    facet_wrap( ~ SASAP.Region,
                scales = "free_y") +
    ylab("Escapement") +
    ggtitle("Annual Salmon Escapement by Region") +
    my_theme +
    theme(axis.text.x = element_text(angle = 45,
                                     vjust = 0.5))

annual_region_plot

# use tab to align code

```
 
 ### Saving plots
```{r saving}

# ggsave by default saves the last plot you saved

ggsave("figs/annual_esc_region.png", annual_region_plot, width = 12, height = 8, units = "in") 

# saved in R Studio in files -> figs 

```
 
 ## Maps with `leaflet`
 
 Make an interactive map that plots all regions
```{r get locations}

#need to locations to make map (creating subset of locations to make map)

location <- escapement %>% 
  distinct(Location, Latitude, Longitude) %>%  #creating a subset to get lat/long of locations
  drop_na()

```
 
 Creating map with `leaflet()`
 
```{r create map}

# leaflet makes interactive maps
# leaflet uses pipe instead of plus

leaflet(location) %>%  # tell leaflet I want to plot locations
  addTiles() %>% 
  addMarkers(lng = ~Longitude,
             lat = ~Latitude,
             popup = ~Location)
  
```
 
 Map with different basemap
 
```{r dif base map}

leaflet(location) %>%
    addWMSTiles(
        "https://www.gebco.net/data_and_products/gebco_web_services/web_map_service/mapserv?request=getmap&service=wms&BBOX=-90,-180,90,360&crs=EPSG:4326&format=image/jpeg&layers=gebco_latest&width=1200&height=600&version=1.3.0", # add specific map
        layers = 'GEBCO_LATEST',
        attribution = "Imagery reproduced from the GEBCO_2022 Grid, WMS 1.3.0 GetMap, www.gebco.net"
    ) %>%
    addCircleMarkers( # change look of markers
        lng = ~ Longitude,
        lat = ~ Latitude,
        popup = ~ Location,
        radius = 5, # size of marker
        # set fill properties
        fillColor = "salmon", # color of marker
        fillOpacity = 1,
        # set stroke properties
        stroke = T,
        weight = 0.5,
        color = "white", #color outside marker
        opacity = 1
    )

# can search for other WMSTiles maps

```
 
 