---
title: "Intro to sf package for spatial data and making maps"
author: "Maya Almaraz"
date: "2023-04-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries
```{r libraries}
library(sf)
library(ggplot2)
library(leaflet)
library(scales)
library(ggmap) # base maps to static maps
library(dplyr)
library(readr)
```

## Upload data
- Download data from [link](https://dev.nceas.ucsb.edu/view/urn:uuid:6f07cb25-a4a1-48e8-95cb-74f532f3ce2d) to computer
- Go to data file
- Click upload
- Select zip file from downloads
- See files pop up in data folder in R Studio

## Explore AK Regions data

```{r read ak regions data}
ak_regions <- sf::read_sf("data/ak_regions_simp.shp") # click tab to find file
```

## Plot map
```{r simple plot}
plot(ak_regions)

# plot not great and needs data wrangling
# check class(ak_regions) in console, see sf and data.frame
```

```{r explore data}
head(ak_regions)
glimpse(ak_regions)

# View(ak_regions) in console and see geometry column with spatial data
```

```{r look at CRS}
st_crs(ak_regions)

# learn it's unprojected and thus not good for maps (may explain map above), need to change
```

## Transform CRS of `ak_regions`
```{r transform crs}
ak_regions_3338 <- ak_regions %>% 
  st_transform(crs = 3338) # pipe so don't need to specify x

# see new object 

st_crs(ak_regions_3338) #looking at transformed data

# we will use three EPSG codes (4326 is default, coords for whole world, 3857 used a lot in big spatial tools like google maps)

```

```{r plot again}
plot(ak_regions_3338) # looks way better

```
## sf and the tidyverse

```{r}
colnames(ak_regions_3338)
```

```{r}
ak_regions_3338 %>% 
  select(region)
```

```{r}
ak_regions_3338 %>% filter(region == "Yukon") # within region column only filter for where the value is Yukon

# ?sf::tidyverse in console
```

## Spatial Joins
- Ask how many people live in each of the AK regions
- Read in the population data from a csv and turn it into an sf object
- Use a spatial join (st_join()) to assign each city to a region
- Use group_by() and summarize() to calculate the total population by region
- Save the spatial object you created using write_sf()

```{r}
pop <- read_csv("data/alaska_population.csv")

# View(pop) and class(pop) in console - don't see geometry column, not spatial data

```

```{r}
# can transform population data into spatial data 

pop_4326 <- st_as_sf(pop, # use to transform
                     coords = c("lng", "lat"), 
                     crs = 4326, # dont have metadata so use default crs
                    remove = FALSE) # takes coord data (lat/lng) and removes cause now have geometry data

# View(pop_4326) in console
                    
```

```{r error example}
# join data using st_join

#pop_joined <- st_join(pop_4326, ak_regions_3338, join = st_within) 
# want to know pops within the regions so use st_within

# see error 'st_crs(x) == st_crs(y) is not TRUE', can only join spatial data where crs' are the same, need another transormation 

```

```{r}
pop_3338 <- st_transform(pop_4326, crs = 3338)

st_crs(pop_3338) # see if changed
```

```{r}
pop_joined <- st_join(pop_3338, ak_regions_3338, join = st_within)

# View(pop_joined) in console, see columns for population
```

```{r group by}
pop_region <- pop_joined %>% 
  as.data.frame() %>%  # don't need to specify because of pipe, turn into data frame
  group_by(region) %>% 
    summarize(total_pop = sum(population)) #create new column called total pop

head(pop_region)

# class(pop_region) shows it's still a data frame and you need it to be a spatial or sf object
```
```{r}
# join ak_region_3338 with pop_region by common key region:
pop_region_3338 <- left_join(ak_regions_3338, pop_region, by = "region") 

#colnames(pop_region_3338) and View(pop_region_3338) in console see that added total pop to table
```

```{r plot by total pop}
# square brackets accessing differnt variables
# plot this object and within it plot this variable total pop

plot(pop_region_3338["total_pop"])
```
```{r}
# plot but group by mgmt area

pop_mgmt_3338 <- pop_region_3338 %>% 
  group_by(mgmt_area) %>% 
  summarize(total_pop = sum(total_pop), do_union = FALSE) #do union within summarize to preseve geometries for alaska regions

head(pop_mgmt_3338)
```
```{r}
#plot total pop by mgmt code

plot(pop_mgmt_3338["total_pop"])
```
```{r}
write_sf(pop_region_3338, "data/ak_regions_population.shp") #save spatial regions and then the file path

# see new shape file added to data and some meta data files added
```

## Visualize with ggplot

```{r}
ggplot(pop_region_3338) +
  geom_sf(aes(fill = total_pop))+ #fill aesthetic with total pop cause that's what we want to show
  labs(fill = "Total Population")+ # change legend title
  scale_fill_continuous(labels = comma, # change from sci notation
                        low = "khaki", # change colors
                        high = "firebrick")+
  theme_bw() # change background

#alteratively

ggplot(pop_region_3338) +
  geom_sf(aes(fill = log10(total_pop)))+ # change scale to log
  labs(fill = "Total Population")+ 
  scale_fill_viridis_c(labels = comma)+ #color blind colors
  theme_bw()

# rainbow parentheses: Code --> rainbow parenthese
#can add scale bar using ggsn
```
## Rivers spatial data

```{r read in rivers}

rivers_3338 <- read_sf("data/ak_rivers_simp.shp") #press tab to find file

# st_crs(rivers) #see it's also alaksa albers, add 3338 to object

# rm(rivers) to remove old object from environment

```

```{r}
# map total pop and added all points of communities/cities:

ggplot()+
  geom_sf(pop_region_3338, mapping=aes(fill = total_pop))+ # need to call the argument mapping
  geom_sf(pop_3338, mapping=aes())+ # looking for and aes
  geom_sf(rivers_3338, mapping=aes())

# two geom_sf because over laying two spatial datasets

# colnames(rivers_3338) in console

# make map look nicer

ggplot()+
  geom_sf(pop_region_3338, mapping=aes(fill = log10(total_pop)))+ # need to call the argument mapping and change to log10
  geom_sf(pop_3338, mapping=aes(), color = "red4", size = 0.5)+ # looking for and aes and change dot color and change size of dots
  geom_sf(rivers_3338, mapping=aes(linewidth = StrOrder))+ #map fewer rivers
  scale_linewidth(range = c(0.05, 0.05), guide = "none")+ # make rivers skinnier and get rid of second legend
  scale_fill_viridis_c(labels = comma)+ # change color and get rid of sci notation
  labs(fill = "Total Population",
       title = "Map of Alaska Populations")+ # add title
  theme(plot.title = element_text(hjust = 0.5)) # center title
  
```
Skip section 15.6

## Leaflet interactive map

```{r}
# create a custom

epsg3338 <- leaflet::leafletCRS(
    crsClass = "L.Proj.CRS",
    code = "EPSG:3338",
    proj4def =  "+proj=aea +lat_1=55 +lat_2=65 +lat_0=50 +lon_0=-154 +x_0=0 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs",
    resolutions = 2 ^ (16:7)
)

# now we have a project we can use in our leaflet map (create an object)
```

```{r}

# have the projected but now need crs

pop_region_4326 <- pop_region_3338 %>% st_transform(crs = 4326)

# created two objects (pieces of spatial data) to create a plot
```

```{r}
m <- leaflet(options = leafletOptions(crs = epsg3338)) %>%
    addPolygons(data = pop_region_4326,
                fillColor = "lightgreen",
                weight = 2) # change size of boundary lines

m  # need to call the object to see map
```

